---
title: "RMD baics"
subtitle: "StatPREP R Workshops 2021"
author: "Kate Kozak"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("mosaic")
library("mosaicData")
```

## Basics of RMD

Creating an r chunk. 

```{r}
2*5
sqrt(4)
log(10)
```


Try executing this chunk by clicking the green *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter* (*ctrl+shift+Return* on Windows). These two actions do the same thing-- they run all the lines of code in a chunk, and show you all the output.

You can also run just one line at a time by putting your cursor on a line and pressing *Cmd+Enter*. Then, you'll just see the output for the one line. 


## Reading in data

```{r}
SleepStudy <- read.csv("../Data/SleepStudy.csv") #Reading in an uploaded dataset
SleepStudy_1 <- read.csv("https://krkozak.github.io/MAT160/SleepStudy.csv") #Reading in a github dataset
```

Using datasets from Packages. Go into MosaicData to find datasets in the package. Then use ?and the name of a dataset to look at the codebook.

```{r}
?Alcohol
```

## "knitted" HTML version

This functionality is supported by an R package called `knitr`, so we often talk about "knitted" versions of RMarkdown and R Notebook documents. 

When you save the notebook, an HTML file containing the code and output will be saved alongside it. This makes a nice, polished report of your work to share.

Click the *Preview* button at the top of this document or press *Cmd+Shift+K* (*Ctrl+Shift+K* on Windows) to preview the HTML file. Try clicking *Preview* now.

To knit a file click the "knit" button at the top of this document.


